---
title: "Preprocess GeoMX data"
author: "Clemens Hug"
date: "2023-10-12"
output: html_document
---

```{r setup}
library(tidyverse)
library(readxl)
library(here)
library(qs)

theme_set(theme_minimal())

synapser::synLogin()
syn <- synExtra::synDownloader("~/data", cache = TRUE)
```

## Load counts

```{r load_counts}
read_count_matrix_excel <- function(path) {
  read_excel(path, "TargetCountMatrix") %>%
    # PanCK sometimes lowercase, make it all uppercase
    rename_with(\(x) str_replace(x, "ck", "CK")) %>%
    arrange(TargetName)
}

count_matrix <- read_count_matrix_excel(
  here("All_batch_combined", "Q3 norm_all_batch.xlsx")
)

sample_qc_metrics <- read_excel(
  here("All_batch_combined", "Q3 norm_all_batch.xlsx"),
  "SegmentProperties"
) %>%
  mutate(
    sample_id = str_replace(SegmentDisplayName, "ck", "CK")
  )
```

## Metadata

Remove any Immune control samples. Not needed here

```{r metadata}
load_metadata <- function(path) {
  read_excel(path) %>%
    mutate(
      sample_id = paste(`Scan name`, `ROI (label)`, `Segment (Name/ Label)`, sep = " | "),
      across(
        c(sample_id, `Segment (Name/ Label)`),
        ~str_replace(.x, "ck", "CK")
      )
    )
}

meta <- load_metadata(
  here("All_batch_combined", "ROI_annotation_combined_all_batch_F.xlsx")
) %>%
  filter(
    !str_detect(Lesions_categories_All, fixed("control", ignore_case = TRUE)),
    !str_detect(Cell_Types_control, fixed("control", ignore_case = TRUE)),
    Cell_Types_control != "Tnecosis"
  ) %>%
  mutate(
    sample_type = case_when(
      `Incidental? Y/N` == "Y" ~ "incidental",
      `Incidental? Y/N` == "N" ~ "cancer",
      TRUE ~ NA_character_
    )
  )
```

Check sample overlap. Verified that missing samples in count matrix were
removed during QC. Also verified that missing samples in metadata are
due to removal of immune control and necrosis samples.

LSP16163 is present in count matrix but not in metadata. This sample passed QC
but was excluded because, according to Tanjina:
> I have to take that sample off because we are not including that sample on cycif.
> That sample only had FT and Fimbriae on geomx slide too (tissue loss). So, you
> may find that sample still exist in count matrix but not in my annotation file.
> Count matrix file basically after normalising all the samples, so its included.
> for my other pathway analysis and Differential gene expression, I am not
> including that sample. So, your segment number should ideally be the same as
> mine

```{r}
meta %>%
  filter(!sample_id %in% colnames(count_matrix)) %>%
  View()

setdiff(colnames(count_matrix), meta$sample_id)

meta %>%
  filter(`Scan name` == "LSP16145_Scan1") %>%
  View()

meta_selected <- meta %>%
  filter(sample_id %in% colnames(count_matrix))

dir.create("data_processed")
write_csv(
  meta_selected,
  "data_processed/meta_selected.csv"
)
```

## Assemble datasets

Count matrix contains both cancer samples from first GeoMX run as well as
incidental samples from second GeoMX run. We want to separate these two for some
analyses and normalize them separately. For other analyses we want a combined
dataset. Creating both separate and combined datasets here.

```{r assemble_data}
count_matrices <- meta_selected %>%
  group_by(sample_type) %>%
  summarize(
    data = list(select(count_matrix, TargetName, any_of(sample_id))),
    .groups = "drop"
  ) %>%
  bind_rows(
    summarize(
      .,
      across(sample_type, \(x) paste(x, collapse = "+")),
      across(data, \(x) list(reduce(x, \(y, z) left_join(y, z, by = "TargetName")))),
      .groups = "drop"
    )
  ) %>%
  mutate(
    data_mat = map(data, column_to_rownames, "TargetName") %>% map(as.matrix)
  )

dir.create("data_processed")
qsave(
  count_matrices,
  "data_processed/count_matrices.qs"
)
```

## Library size correction

Don't do library size correction anymore because we are starting with Q3
normalized data from GeoMX.

Also storing log-transformed and scaled (0 mean, 1 SD) versions of the
count matrices.

```{r}
count_matrices_trans <- count_matrices %>%
  select(sample_type, data_mat) %>%
  crossing(transform = c("raw", "log")) %>%
  mutate(
    data_mat = map2(
      data_mat, transform,
      \(x, y) {
        if (y == "raw") {
          x
        } else {
          log10(x)
        }
      }
    )
  ) %>%
  crossing(scaled = c("unscaled", "scaled")) %>%
  mutate(
    data_mat = map2(
      data_mat, scaled,
      \(x, y) {
        if (y == "unscaled") {
          x
        } else {
          t(scale(t(x), center = TRUE, scale = TRUE))
        }
      }
    )
  )

count_matrices_trans$data_mat[[8]] %>%
  as_tibble(rownames = "TargetName") %>%
  pivot_longer(
    -TargetName,
    names_to = "sample_id",
    values_to = "count"
  ) %>%
  filter(is.na(count))

count_matrices_trans_long <- count_matrices_trans %>%
  mutate(
    data = map(
      data_mat,
      \(x) as_tibble(x, rownames = "TargetName") %>%
        pivot_longer(
          -TargetName,
           names_to = "sample_id",
          values_to = "count"
        )
    )
  ) %>%
  select(-data_mat) %>%
  unnest(data) %>%
  pivot_wider(
    names_from = c(transform, scaled),
    values_from = count
  )

write_csv(
  count_matrices_trans_long,
  "data_processed/count_matrix_all_trans.csv.gz"
)
qsave(
  count_matrices_trans,
  "data_processed/count_matrix_all_trans.qs"
)
# count_matrices_trans <- qread("data_processed/count_matrix_all_trans.qs")
```

## Gene sets

We selected gene sets from the literature that are associated with different
aspects of cancer progression and immunity. We want to check if their expression
changes during cancer progression.

Gene sets are grouped by relatedness.

```{r load_gene_sets}
gene_set_file <- here("All_batch_combined", "Gene_signatures_F_08222023_short.xlsx")

gene_sets <- map(
  excel_sheets(gene_set_file) %>%
    # Remove sheet with references
    setdiff(c("references")) %>%
    set_names(),
  ~read_excel(gene_set_file, .x) %>%
    pivot_longer(everything(), names_to = "gene_set", values_to = "gene") %>%
    drop_na() %>%
    mutate(
      gene_unique = paste(gene, gene_set, sep = "_")
    )
) %>%
  bind_rows(.id = "gene_set_group")

gene_sets %>%
  dplyr::count(gene) %>%
  arrange(desc(n))

write_csv(
  gene_sets,
  "data_processed/gene_sets.csv"
)
```

```{r}
setdiff(gene_sets$gene, count_matrix$TargetName)
```

14 genes in gene sets not in count matrix. Filtered out during QC

```{r}
gene_sets_selected <- gene_sets %>%
  filter(
    gene %in% count_matrix$TargetName
  )

gene_sets_unique <- gene_sets_selected %>%
  group_by(gene_set) %>%
  arrange(gene) %>%
  summarize(
    gene_str = paste(gene, collapse = "|"),
    genes = list(gene),
    .groups = "drop"
  ) %>%
  # Keep only one unique gene set. If multiple gene sets have the same genes,
  # keep only one
  group_by(gene_str) %>%
  slice_head(n = 1) %>%
  ungroup()

qsave(
  gene_sets_unique,
  "data_processed/gene_sets_unique.qs"
)
```

## Check gene expression

```{r}
expression_plot_data <- count_matrices_trans_long %>%
  inner_join(
    gene_sets_selected %>%
      distinct(gene),
    by = c("TargetName" = "gene")
  )

expression_plot_data %>%
  group_by(TargetName) %>%
  summarize(
    q3 = quantile(raw_unscaled, probs = .75),
    .groups = "drop"
  ) %>%
  arrange(q3)
# MIN_COUNT <- min(expression_plot_data$raw_unscaled[expression_plot_data$raw_unscaled > 0]) * .5

p <- expression_plot_data %>%
  mutate(
    TargetName = factor(TargetName, levels = {group_by(., TargetName) %>%
      summarize(across(raw_unscaled, mean)) %>%
      arrange(raw_unscaled) %>%
      pull(TargetName)})
  ) %>%
  filter(TargetName %in% head(levels(TargetName), n = 50)) %>%
  ggplot(
    aes(TargetName, raw_unscaled)
  ) +
  geom_violin() +
  scale_y_log10(breaks = c(c(1, 2, 5, 10, 20, 50, 100))) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  )

```

## QC metrics

```{r}
write_csv(
  sample_qc_metrics,
  "data_processed/qc_metrics.csv.gz"
)
```


## Store in Synapse

```{r upload_synapse}
synExtra::synStoreMany(
  c(
    "data_processed/count_matrices.qs",
    "data_processed/gene_sets.csv",
    "data_processed/gene_sets_unique.qs",
    "data_processed/meta_selected.csv",
    "data_processed/count_matrix_all_trans.csv.gz",
    "data_processed/count_matrix_all_trans.qs",
    "data_processed/qc_metrics.csv.gz"
  ),
  parentId = "syn52658424",
  used = "https://www.dropbox.com/sh/9qj3elozuupzemo/AADs33pqqzUFHgttpf_7RyEYa?dl=0",
  executed = "https://github.com/clemenshug/pre-ovarian-cancer-atlas/blob/main/preprocess_data.Rmd",
  forceVersion = FALSE
)
```
